### 2022-10-28 Manuscripts/deNovoRates.tex
modified:   Modeling/DeNovoProb.jl

Next, we tried to find out if some amino acid substitutions are more probable, than the others. To this end, we calculated the gain probability ($G$) for every amino acid. Specifically for every amino acid $i$, we calculated the average probability that it substitutes any of the other nineteen amino acids, due to mutations. More precisely, the total probability that an amino acid $j$ will mutate to an amino acid $i$ is the product of the expected frequency of amino acid $j$ ($P_j$), and the substitution rate from $i$ to $j$ ($\mu_{j\to i}$). The gain probability for amino acid $i$ is thus defined as:

$$G_i = \sum_{j \neq i} P_j \times \mu_{j\to i}$$

@@

$$L_i = \sum_{j \neq i} \mu_{i\to j}$$

@@

>>>>

Next, we tried to find out if some amino acid substitutions are more probable, than the others. To this end, we calculated the gain probability ($G$) for every amino acid. Specifically for every amino acid $a$, we calculated the average probability that it substitutes any of the other nineteen amino acids, due to mutations. More precisely, the total probability that an amino acid $b$ will mutate to an amino acid $a$ is the product of the expected frequency of amino acid $b$ ($P_b$), and the substitution rate from $a$ to $b$ ($\mu_{b\to a}$). The gain probability for amino acid $a$ is thus defined as:

$$G_a = \sum_{b \neq a} P_b \times \mu_{b\to a}$$

@@

$$L_a = \sum_{b \neq a} \mu_{a\to b}$$

@@

Next, we investigated if some amino acid substitutions are more common than the others. As expected, amino acid substitutions that require more than one nucleotide change were much less probable than those needing just a single nucleotide change. Therefore, we focused our analysis on amino acid substitutions whose total probabilities were more than that of a double nucleotide mutation (note that this procedure identifies amino acid substitutions that are possible via single nucleotide mutation, but it does not exclude multinucleotide mutations from the calculation of amino acid substitution probabilities). Although double nucleotide mutations are less probable than single mutations, the mutation bias makes causes different single mutations to occur with different probabilities. Thus it is very likely that an amino acid substitution (for example K$\to$E) may not be as probable as the reverse substitution (E$\to$K). To identify if such a directionality exists, we calculated the log transformed ratio of forward and reverse substitution probability, $D_{ab}$, for every pair of amino acids, $a$ and $b$. A positive value of $D_{ab}$ means that $a$ to $b$ substitution is more likely than $B$ to $a$ substitution, and \textit{vice versa}.

$$ \textit{D}_{ab} = \log_2\left(\frac{\mu_{a\to b}}{\mu_{b\to a}}\right)$$

We restricted this analysis to substitutions that can be achieved via single nucleotide mutations. We found that most substitutions indeed had a preferred direction $\textit{D}_{ab} \neq 0$. The median absolute value of $\textit{D}_{ab}$ was 0.905 and the maximum value was 2.836. This means that for 50\% of the pairs we analysed, one substitution direction was 1.87 times more probable than the other. For example, E$\to$K is 1.8 times more probable than K$\to$E. This may appear as a small number but many such directional substitutions occurring in an evolving protein could indicate a directional selection.
